Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-24 paper.tex
Work in plink2, some additional compression refs

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
Large genome variation datasets are also being generated for other organisms
and a variety of purposes including 
agriculture~\citep{ros2020accuracy,wang2023rice}, 
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats 
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits 
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

@@

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
in here.]
Methods have also been developed which store variation data 
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or require storage larger 
than uncompressed VCF~\citep{al2023critical}.

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
Large genome variation datasets are also being generated for other organisms
and a variety of purposes including 
agriculture~\citep{ros2020accuracy,wang2023rice}, 
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats 
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN formats are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits 
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

@@

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{qiao2012handling, % SpeedGene
deorowicz2013genome, %TGC
sambo2014compression, % snpack
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
in here.]
The widely-used PLINK binary format stores genotypes in a 
packed binary representation, supporting only biallelic
variants without phase information.
The PLINK 2 PGEN format~\citep{pgen2024} is more general
and compact than PLINK, compressing data using specialised
genotype compression methods~\cite{sambo2014compression}.
Methods have also been developed which store variation data 
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or have storage requirements larger
than uncompressed VCF~\citep{al2023critical}.
